Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-10-18 paper/ms.tex

\subsection{Derivative of likelihood with respect to Kernel parameters}

The derivative of the likelihood can be computed in $O(N)$ operations as follows.

The derivative of the log likelihood with respect to model parameters ${\bf x}$ is given by:
\begin{equation}
    \frac{\partial \ln{\cal L}}{\partial x_i} = (\ln{\cal L})^\prime = -\frac{1}{2}\sum_i \frac{K_{ii}^\prime}{K_{ii}}
    - \frac{1}{2} {\bf y}^T {\bf K}^{-1} {\bf K}^\prime {\bf K}^{-1} {\bf y},
\end{equation}
where $^\prime$ means the derivative wrt $x_i$.
The kernel may be written as:
\begin{equation}
    K_{ij}({\bf \alpha},{\bf \beta}) = \left\{
    \begin{array}{cc}
       w_i + \sum_{l=1}^p \alpha_l & j = i \\
       \sum_{l=1}^p \alpha_l \exp{(-\beta_l |t_i-t_j|)}  & j \ne i
    \end{array}
    \right.
\end{equation}
Since this equation is the sum of $p$ semi-separable matrices, then applying the product rule to compute the derivative will give a single semi-separable matrix for the derivative with respect to each kernel parameter.

Now, letting ${\bf x} = (w,{\bf \alpha},{\bf \beta})$, then the derivative may be written as:
\begin{eqnarray}
    \frac{\partial K_{ii}}{\partial w} &=& \delta_{ij}\\
    \frac{\partial K_{ij}}{\partial \alpha_k} &=&
     \left\{
    \begin{array}{cc}
       1 & j = i \\
        \exp{(-\beta_k |t_i-t_j|)}  & j \ne i
    \end{array}
    \right.\\
    \frac{\partial K_{ij}}{\partial \beta_k} &=&
     \left\{
    \begin{array}{cc}
       0  & j = i \\
        -\alpha_k|t_i-t_j|\exp{(-\beta_k |t_i-t_j|)}  & j \ne i
    \end{array}
    \right.
\end{eqnarray}
The first term in the derivative of the likelihood can be computed in $O(N)$
evaluations, while the second term can be computed in three steps:
\begin{itemize}
\item First, solve ${\bf z_1} = {\bf K}^{-1}{\bf y}$ with the GRP solver.
\item Next, create an e
\end{itemize}
Writing this term out:

>>>>

\section{Generalized Press-Rybicki}

The general problem to be solved for Gaussian Process analysis of stellar time series is to evaluate
the likelihood function,
\begin{equation}
{\cal{L}} = \ln{p(\mathbf{y}\vert {\mathbf{t},\bvec{\theta}},\bvec{\alpha})} = -\frac{1}{2}\ln{\vert \mathbf{K}(\bvec{\alpha}) \vert}
- \frac{N_{time}}{2} \ln{(2\pi)}
    - \frac{1}{2} {\mathbf r} (\bvec{\theta} ) ^T \mathbf{K}(\bvec{\alpha})^{-1} \mathbf{r}(\bvec{\theta}),
\end{equation}
where $\mathbf{y}$ are the measured time series (for example fluxes or radial velocities) at times
$\mathbf{t}$, while the data are modeled with a function $m(\mathbf{t};\bvec{\theta})$ (for example,
a transit model or Keplerian orbital model) which leaves residuals $\mathbf{r}(\bvec{\theta}) =
\mathbf{y} - m(\mathbf{t};\bvec{\theta})$. % Note: I'm using the same notation as the peerless project. -EA
The vector $\bvec{\alpha}$ specifies the parameters of the kernel, while the vector $\bvec{\theta}$
specifies the parameters of the physical model.  In addition to this likelihood, a prior may be placed on 
the values of these model parameters (either the physical or kernel); the log of this prior may be
added to this log likelihood function.  Note that the kernel
parameters may also have a physical meaning, but they describe a noisy process, and thus affect
the correlated variability of stellar noise rather than a deterministic model as encapsulated in
$m(\mathbf{t};\bvec{\theta})$.

We express the kernel as the sum of exponential functions...

%\subsection{Derivative of likelihood with respect to Kernel parameters}
%
%The derivative of the likelihood can be computed in $O(N)$ operations as follows.
%
%The derivative of the log likelihood with respect to model parameters ${\bf x}$ is given by:
%\begin{equation}
%    \frac{\partial \ln{\cal L}}{\partial x_i} = (\ln{\cal L})^\prime = -\frac{1}{2}\sum_i \frac{K_{ii}^\prime}{K_{ii}}
%    - \frac{1}{2} {\bf y}^T {\bf K}^{-1} {\bf K}^\prime {\bf K}^{-1} {\bf y},
%\end{equation}
%where $^\prime$ means the derivative wrt $x_i$.
%The kernel may be written as:
%\begin{equation}
%    K_{ij}({\bf \alpha},{\bf \beta}) = \left\{
%    \begin{array}{cc}
%       w_i + \sum_{l=1}^p \alpha_l & j = i \\
%       \sum_{l=1}^p \alpha_l \exp{(-\beta_l |t_i-t_j|)}  & j \ne i
%    \end{array}
%    \right.
%\end{equation}
%Since this equation is the sum of $p$ semi-separable matrices, then applying the product rule to compute the derivative will give a single semi-separable matrix for the derivative with respect to each kernel parameter.
%
%Now, letting ${\bf x} = (w,{\bf \alpha},{\bf \beta})$, then the derivative may be written as:
%\begin{eqnarray}
%    \frac{\partial K_{ii}}{\partial w} &=& \delta_{ij}\\
%    \frac{\partial K_{ij}}{\partial \alpha_k} &=&
%     \left\{
%    \begin{array}{cc}
%       1 & j = i \\
%        \exp{(-\beta_k |t_i-t_j|)}  & j \ne i
%    \end{array}
%    \right.\\
%    \frac{\partial K_{ij}}{\partial \beta_k} &=&
%     \left\{
%    \begin{array}{cc}
%       0  & j = i \\
%        -\alpha_k|t_i-t_j|\exp{(-\beta_k |t_i-t_j|)}  & j \ne i
%    \end{array}
%    \right.
%\end{eqnarray}
%The first term in the derivative of the likelihood can be computed in $O(N)$
%evaluations, while the second term can be computed in three steps:
%\begin{itemize}
%\item First, solve ${\bf z_1} = {\bf K}^{-1}{\bf y}$ with the GRP solver.
%\item Next, create an e
%\end{itemize}
%Writing this term out:
